---
title: "How does education level relate to suicide rate?"
author: "Jiaqi Li and Mackie Zhou"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
---

# Background

According to the National Center for Health Statistics(NCHS), the suicide rate increased 33 percent from 1999 through 2017^[Suicide Mortality in the United States, 1999–2017, https://www.cdc.gov/nchs/products/databriefs/db330.htm], from 10.5 to 14 suicides per 100,000 people. Suicide ranks as the fourth leading cause of death for people ages 35 to 54, and the second for 10 to 34-year-olds. Suicide is a serious social problem. We seek to understand the possible risk factors that contribute to suicide rate by analyzing the relationship between education attainment and suicide rate. Given that socioeconomic status is associated with suicide attempt, __we hypothesize that education attainment, as an important measurement of socioeconomic class, is negatively correlated with suicide rate__.


```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Set R chunk default options:
# -do not output all R messages in html report
# -do not output all R warnings in html report
# -set default figure widths and heights for all code blocks, unless you 
#  override them
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE, fig.width = 16/2, fig.height = 9/2
)
```

```{r, echo = FALSE}
library(tidyverse)
library(sf)
library(maps)
library(USAboundaries)
library(plotly)
library(widgetframe)
library(scales)

# read the ecudational attainment df into R
Edu_complete <- read_csv("Estimate_Educational_attainment_2017.csv")

# read the suicide df into R
suicide_by_state <- read_csv("Suicide_Mortality_by_State.csv")

# read the 50 state names&codes df into R
State_names <- read_csv("50State_names_and_codes.csv")

# US_map
US_map<-USAboundaries::us_states(resolution = "high") %>%
  mutate(lon = purrr::map_dbl(geometry, ~sf::st_centroid(.x)[[1]]),
         lat = purrr::map_dbl(geometry, ~sf::st_centroid(.x)[[2]]))

```


```{r, echo = FALSE}
# select only data in 2017 and optimize the df
suicide_2017 <- suicide_by_state %>% 
  filter(YEAR=="2017") %>% 
  left_join(State_names, by = c("STATE" = "Code")) %>% 
  rename(CODE = STATE,
         STATE = State) %>%
  arrange( RATE ) %>% 
  select(STATE, CODE, DEATHS, RATE)

edu_score_suicide <- suicide_2017 %>% 
  inner_join(Edu_complete, by = c("STATE" = "State")) %>% 
  select(-Id, -Id2) %>% 
  rename(Death_Rate100 = RATE, Education_Level = score) %>% 
  mutate(Death_Rate = Death_Rate100/100) %>% 
  select(STATE, Death_Rate, Education_Level, everything())

scatter <- ggplot(edu_score_suicide, aes(x = Education_Level, y = Death_Rate)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = TRUE) +
  scale_y_continuous(labels = percent) +
  labs(x = "Education Level",
       y = "Suicide Death Rate",
       title = "Education Level Score and Suicide Death Rate of 50 US States in 2017")
scatter
```

# Conclusion & Significance

__Moderate negative correlation__ is found between education attainment and suicide rate, which means that the higher the education attainment, the lower the suicide rate. Possible explanation for this result could be that people who do not drop out of school tend to have better family background, thus they are less likely to have childhood trauma (abusive parents), and are more likely to commit suicide. The result emphasizes the importance of social determinants in suicide risk. 

# Citation